#440 Reproducibility and Statistics Assignment Homework

<!-- This is an Rmd comment. -->
<!--
You can make comments several lines long by inserting
the text in between the the start and the end of the 
comment.

Comments are where I'll tell you what to do. You should
use the space directly below the comments to do what
the comment asks. You should also describe what you're
doing to make it obvious to the reader so they can
easily follow your logic in a fully reproducible way.
For any statistical tests you run, you should explain
the results in plain English and present the statistics
as you would in an APA style manuscript. For any symbols
you don't know how to use/can't use, feel free to use the name
instead (e.g. alpha)
--> 

<!-- Load the politics.csv data file. -->
First we need to load the data. 

```{r}
pols<-read.csv("politics.csv") 
```



<!-- 
Check the structure.
subject, party, testtime, minwage, and sex need to be factors.
optimismscore and income should be integer and numeric vectors.
Fix any variables that aren't right.
-->
Let's take a look at the data.
```{r}
str(pols)
```
```
## 'data.frame':    132 obs. of  7 variables:
##  $ subject      : int  1 2 3 4 5 6 7 8 9 10 ...
##  $ party        : Factor w/ 3 levels "democrat","independent",..: 3 3 2 2 2 3 3 2 3 2 ...
##  $ testtime     : Factor w/ 2 levels "post","pre": 2 2 2 2 2 2 2 2 2 2 ...
##  $ optimismscore: int  52 51 69 51 61 31 57 48 42 64 ...
##  $ minwage      : Factor w/ 2 levels "no","yes": 1 1 2 1 2 1 1 1 1 1 ...
##  $ sex          : Factor w/ 2 levels "female","male": 2 2 1 2 2 2 2 2 2 2 ...
##  $ income       : num  37.3 42.3 73 33.8 57.3 ...
```
```{r}
pols$subject<-factor(pols$subject)
```
```{r}
pols$testtime<-factor(pols$testtime, levels=c("pre", "post"))
```
```{r}
str(pols)
```
```
data.frame':  132 obs. of  7 variables:
 $ subject      : Factor w/ 66 levels "1","2","3","4",..: 1 2 3 4 5 6 7 8 9 10 ...
 $ party        : Factor w/ 3 levels "democrat","independent",..: 3 3 2 2 2 3 3 2 3 2 ...
 $ testtime     : Factor w/ 2 levels "pre","post": 1 1 1 1 1 1 1 1 1 1 ...
 $ optimismscore: int  52 51 69 51 61 31 57 48 42 64 ...
 $ minwage      : Factor w/ 2 levels "no","yes": 1 1 2 1 2 1 1 1 1 1 ...
 $ sex          : Factor w/ 2 levels "female","male": 2 2 1 2 2 2 2 2 2 2 ...
 $ income       : num  37.3 42.3 73 33.8 57.3 ...
```

<!-- Find the minimum, mean, and variance for posttest
optimism scores. -->
I need to find the minimum, mean, and variance for posttest optimism scores. I can do this with one command.

```{r}
summary(pols$income)
```
```
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  5.672  27.590  41.220  43.040  56.010 114.800 
```

<!-- Create a histogram of posttest optimism scores.
Feel free to use the default ugly settings since
we didn't cover how to do this in ggplot2. If
you're a real go-getter, see if you can figure
out how to make a prettier histogram with better
labels, etc. using ggplot2. -->
In order to see how the incomes are distributed, let's create a histogram using only the 
posttest data.

```{r}
hist(pols$income[pols$testtime=="post"])
```

<!-- Run a test to determine whether political affiliation
is independent of one's gender.
Present the data supporting your analyses in graphical or
tabular form.
-->
Let's see if political affiliation is independent of one's gender. I need to calaculate the frequency in which the genders appear in different political groups. I used only the pretest data so they were not counted twice.
Then a Chi-Square test will be used to test the hypothesis that affiliation 
and gender are independent. 

```{r}
tab<-table(pols$party[pols$testtime=="pre"],pols$sex[pols$testtime=="pre"])
tab
```

```
             female male
  democrat        14   12
  independent      7   10
  republican      12   11
```

These are the frequencies both genders fell into the parties. 
```{r}
chisq.test(pols$party[pols$testtime=="pre"],pols$sex[pols$testtime=="pre"])
```
```
Pearson's Chi-squared test

data:  pols$party[pols$testtime == "pre"] and pols$sex[pols$testtime == "pre"]
X-squared = 0.72674, df = 2, p-value = 0.6953
```

```
Political affiliation is independent of gender, 
Chi-Square(2)=0.73, p=.70
```

<!-- Run a test to determine whether or not males and females
have different incomes.
Create a graph to display your data that includes the relevant
means and standard errors.-->

To determine if males and females have different incomes we will run an 
independent t-test.

```{r}
t.test(pols$income[pols$sex=="male" & pols$testtime=="pre"],pols$income[pols$sex=="female" & pols$testtime=="pre"])
```

```
Males and females do differ in income.
t(61.6)= -1.57, p= .12
```

To create the figure, I need to include the dplyr library. Then
I'll use it to summarize the politics data set and find the means
and standard errors for the different conditions. Then I'll need
the ggplot2 library to create a nice plot that's easy to work with.
I'll do the same thing for any other graphs I create unless I
specify otherwise.
```{r}
library("dplyr")
```
```
Attaching package: ‘dplyr’

The following object is masked from ‘package:stats’:

    filter

The following objects are masked from ‘package:base’:

    intersect, setdiff, setequal, union
```
```{r}
temp<-pols[pols$testtime=="pre",]%>%group_by(sex)%>%summarize(means=mean(income),sems=sd(optimismscore)/sqrt(length(optimismscore)))
library("ggplot2")
f<-ggplot(temp,aes(x=factor(sex),y=means))+geom_bar(stat="identity", color="black",fill=c("deeppink","deepskyblue2"))+geom_errorbar(aes(ymax=means+sems, ymin=means-sems), width=.1)
f
```

This is a bare minimum graph, let's create a better one.  

```{r}
f<-f+ggtitle("Inocme by Sex")+
    labs(x="Sex", y="Income\n(higher=more income")+
    scale_x_discrete(breaks=c("female","male"),labels=c("Female","Male"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(), axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())
f
```


<!-- See if male and females with different political affiliations
have different posttest optimism scores. Create a graph to display
your data that includes the relevant means and standard errors.-->

In order to determine if females and males with different political affiliations have different posttest optimism scores we need to perform and ANOVA. 

```{r}
summary(aov(optimismscore~party*sex,data=pols[pols$testtime=="post",]))
```

```
Optimisim score is related to political affliations, F(2,60)= 27.06, p= 4.2e-09, but 
not to sex F(1,60)= .04, p= .84, or an interaction between affiliation and gender, F(2
,60)=1.214, p=.304. 
```

To create the plot and manually adjust the fill, I need to use
hexadecimal color codes. To convert the named colors I like
to hexadecimal, I need to use the gplots library.

```{r}
temp<-pols[pols$testtime=="post",]%>%group_by(party,sex)%>%summarize(means=mean(optimismscore),sems=sd(optimismscore)/sqrt(length(optimismscore)))
library("gplots")
col1=col2hex("deeppink")
col2=col2hex("deepskyblue2")
f<-ggplot(temp, aes(x=party, y=means, fill=sex))+
    geom_bar(stat="identity",position=position_dodge())+
    scale_fill_manual(values=c(col1,col2),name="Sex",breaks=c("female","male"),labels=c("Female", "Male"))+
    theme(legend.key=element_rect(color="black"))+
    geom_errorbar(aes(ymax=means+sems, ymin=means-sems),width=.2,position=position_dodge(.9))+
    ggtitle("Optimismscore by Sex and Political Affiliation")+
    labs(x="Political Party Affiliation",y="Optimismscore\n(tens)")+
    scale_x_discrete(breaks=c("democrat","independent","republican"),labels=c("Democrat","Independent","Republican"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(),axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())+
    theme(legend.position=c(.2,.76))+
    theme(legend.background=element_blank())+
    theme(legend.background=element_rect(color="black"))+
    theme(legend.title=element_blank())+
    theme(legend.title=element_text(size=12))+
    theme(legend.title.align=.5)+
    theme(legend.text=element_text(size=10,face="bold"))
f
```


<!-- Create a line graph for the same data. With
error bars of course!-->

Now we need to creat a line graph for the same data used above. 

```{r}
f<-ggplot(temp, aes(x=party, y=means, group=sex, color=sex))+
    geom_line(size=1)+
    geom_point(size=2)+
    scale_color_manual(values=c(col1,col2),name="Sex",breaks=c("female","male"),labels=c("Female", "Male"))+
    geom_errorbar(aes(ymax=means+sems, ymin=means-sems),width=.2)+
    ggtitle("Optimismscore by Sex and Political Affiliation")+
    labs(x="Political Party Affiliation",y="Optimismscore\n(tens)")+
    scale_x_discrete(breaks=c("democrat","independent","republican"),labels=c("Democrat","Independent","Republican"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(),axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())+
    theme(legend.position=c(.2,.76))+
    theme(legend.background=element_blank())+
    theme(legend.background=element_rect(color="black"))+
    theme(legend.title=element_blank())+
    theme(legend.title=element_text(size=12))+
    theme(legend.title.align=.5)+
    theme(legend.text=element_text(size=10,face="bold"))
f
```

<!-- Run a test to see if optimism scores pre- and post
watching videos depends on sex.-->

To see if optimism scores before and after watching videos
vary depending on sex, I'm going to run a
2-way mixed ANOVA. I'm using a mixed ANOVA because the same
people indicated their optimism before and after the videos
(i.e. a within-subjects factor) and because different
people are affiliated with different political parties (i.e.
a between-subjects factor).

```{r}
summary(aov(optimismscore~testtime*sex+Error(subject/testtime),data=pols))
```

```
Error: subject
          Df Sum Sq Mean Sq F value Pr(>F)
sex        1     80    80.4   0.119  0.731
Residuals 64  43105   673.5               

Error: subject:testtime
             Df Sum Sq Mean Sq F value   Pr(>F)    
testtime      1  770.9   770.9  41.299 1.87e-08 ***
testtime:sex  1    0.9     0.9   0.049    0.825    
Residuals    64 1194.7    18.7                     
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```



Optimism scores were not related to sex, F(1,64)= .049, p= .825, but was affected by watching videos, F(1,64)= 41.3, p<.01.  There was no interaction between taking the test before or after watching videos and their party affiliation.  



<!-- Run a Multiple Regression Analysis to see
whether you can predict someones posttest optimism
scores from their pretest optimism scores and party
affiliation. Create a figure that shows the distributions
of pre- and posttest optimism scores and sex and create
lines that best fit the data for both sexes regardless of
whether the lines are warranted based on the statistical
analyses-->

I'm going to perform Multiple Regression
to predict posttest optimism scores from their pretest optimism scores and party 
affiliation.  



